Statistical analysis of periodic data in neuroscience

### 2021-06-15 manuscript/manuscript.Rmd
Revisions in response to reviewer comments

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# reasonably compact code to check which packages are installed, 
# install the missing ones, and activate all

packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','devtools')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

nsims <- 1000  # global value for the whole script - increase to 100000 for final run, but note this takes around 16 hours to complete (depending on processor speed)

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase), either to test for the presence of a signal, or to compare signals across different conditions. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here the performance of two multivariate extensions of the T-test are compared: Hotelling's $T^2$ and a variant called $T^2_{circ}$. A novel test of the assumptions of $T^2_{circ}$ is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed. The $T^2_{circ}$ statistic is then extended to multi-level designs, resulting in a new statistical test termed $ANOVA^2_{circ}$. This has identical assumptions to $T^2_{circ}$, and is shown to be more sensitive than MANOVA when these assumptions are met. The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run. Implementations of these novel tools are provided as an _R_ package and a _Matlab_ toolbox, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

A widely used paradigm in many branches of neuroscience is to drive the nervous system using periodic stimuli. This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations. These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded using invasive methods from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987], or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982]. Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014]. In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position. Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].

A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal. The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response (see Figure \@ref(fig:fourierexplain)a,b). Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \@ref(fig:fourierexplain)c), or equivalently as complex numbers with real and imaginary components (Figure \@ref(fig:fourierexplain)d). In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data (or equivalently power or signal-to-noise ratios) only [e.g. @Liu-Shuang2014; @Afsari2014; @Itthipuripat2014; @McFadden2014; @Vanegas2015; @Smith2017; @Hou2020] (note this is a non-exhaustive selection of recent studies). Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if signals differ between two (or more) experimental conditions. However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below. An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components). Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).

@@

@@

\begin{equation}
\label{eq:t2c}
T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2}
\end{equation}

@@

In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms. It is first demonstrated why parametric univariate statistics are inappropriate for such data, because amplitudes for weak signals are not normally distributed. Then conditions are investigated under which either the $T^2$ or $T^2_{circ}$ statistic should be used. The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified. A novel method for testing the assumptions of the $T^2_{circ}$ statistic is developed, based on calculating the condition index of a multivariate data set. Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed. Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA. Finally, the proposed techniques are demonstrated on two example data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.

@@

@@

@@

Typical solutions for dealing with skew, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, and should be used in preference to univariate methods.

# Conditions under which $T^2_{circ}$ is more sensitive than $T^2$

Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than univariate T-tests, as well as avoiding problems caused by using absolute amplitude values. Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$, which is expected on theoretical grounds because the additional degrees of freedom in the $T^2_{circ}$ test make it more efficient. Their simulations involved generating random data sets of different sample sizes, and different signal strengths (sample means), and comparing the number of such tests where each statistic was significant. The simulations showed the largest advantage for $T^2_{circ}$ for effect sizes around _d_ = 1 (where the sample mean is equal to the sample standard deviation). The advantage appeared to be stronger for smaller sample sizes.

Here these simulations are replicated and extended (see Figure \@ref(fig:powerfig)), and it is shown that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8). These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use can be restricted to small sample studies.

```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition). Panels a-c replicate conditions reported by Victor and Mast (1991). Panels (d) and (e) show a wider range of conditions for each statistic. Panel (f) shows the difference between the two statistics, with contour lines indicating differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}

layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
abc <- c('a','b','c')

@@

tsqmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
circmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
for (n in 1:nsims){
  for (s in 1:length(samplesizes)){
    for (s2 in 1:length(dlist)){
    data <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
    data[,1] <- data[,1] + dlist[s2]
    a <- tsqh.test(data)
    if (a[5]<0.05){tsqmatrix[s,s2] <- tsqmatrix[s,s2] + 1}
    a <- tsqc.test(data)
    if (a[5]<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}
  }}
}

tsqpropM <- tsqmatrix/nsims
circpropM <- circmatrix/nsims
tsqvscirc <- circpropM - tsqpropM
tsqvscirc <- pmax(tsqvscirc,0)  # sets any negative values to 0

@@

for (n in 1:3){
par(pty="s")
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2.2, cex.lab=1.5)
title(paste('d =',dlist[vinds[n]]),cex.main=2)
lines(actualslog,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
lines(actualslog,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])

points(actualslog,tsqpropM[,vinds[n]],lwd=3,pch=15,cex=1.5,col=nbdtpal[1])
points(actualslog,circpropM[,vinds[n]],lwd=3,pch=16,cex=1.5,col=nbdtpal[2])

text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)

if (n==1){legend(3.8,1,c(expression(T^2),expression('T'[circ]^2)),pch=15:16,cex=1.5,col=nbdtpal[1:2],box.lwd=2)}
}

par(pty="s")
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
ticklabelsx <- 2^ticklocsx        # set labels for x ticks
ticklabelsy <- seq(0,4,1)    # set labels for y ticks
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(expression(T^2),cex.main=2)
image(slist,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
text(1.1,3.8,'(d)',cex=2)
legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Power')

@@

@@

@@

Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on satisfying the $T^2_{circ}$ test's more stringent assumptions. The two variables must be independent (i.e. uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (horizontal dashed lines). However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the sample variances of the two dependent variables grow more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2. In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).

@@

@@

@@

@@

If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers. The Mahalanobis distance [@Mahalanobis1936] is a useful metric for identifying such multivariate outliers so that they can be excluded. It calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points. This means that any correlations in the data set are taken into account when calculating the distance metric, _D_.

@@

@@

@@

where $x_{i,k}$ is the $i$th data point from group _k_, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as $df_R = 2((Nk)-k)$, where _N_ is the number of observations per group, and _k_ is the number of groups. A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom.

@@

layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))
abc <- c('a','b','c')

@@

text(1.1,0.95,paste('(',abc[n],')',sep=''),cex=2)

@@

@@

pos <- coordinates(c(1,2,4))
plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

@@

@@

@@

@@

allCI <- NULL
allCIp <- NULL
for (cond in 1:7){
  xvals <- Re(cleandata[,cond])/5
  yvals <- Im(cleandata[,cond])/5
  ellRot <- getel(data.frame(xvals,yvals))
  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)

>>>>

```{r setup, include=FALSE}

nsims <- 100000  # global value for the whole script - increase to 100000 for final run, but note this takes around 24 hours to complete (depending on processor speed)
buildfigs <- 1

# reasonably compact code to check which packages are installed, 
# install the missing ones, and activate all
packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','devtools','knitr')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

knitr::opts_chunk$set(echo = TRUE)

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded using a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components (ignoring phase), either to test for the presence of a signal, or to compare signals across different conditions. However, the assumptions of these tests will sometimes be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here the performance of two multivariate extensions of the T-test are compared: Hotelling's $T^2$ and a variant called $T^2_{circ}$. A novel test of the assumptions of $T^2_{circ}$ is developed, based on the condition index of the data (the square root of the ratio of eigenvalues of a bounding ellipse), and a heuristic for excluding outliers using the Mahalanobis distance is proposed. The $T^2_{circ}$ statistic is then extended to multi-level designs, resulting in a new statistical test termed $ANOVA^2_{circ}$. This has identical assumptions to $T^2_{circ}$, and is shown to be more sensitive than MANOVA when these assumptions are met. The use of these tests is demonstrated for two publicly available empirical data sets, and practical guidance is suggested for choosing which test to run. Implementations of these novel tools are provided as an _R_ package and a _Matlab_ toolbox, in the hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

A widely used approach in many branches of neuroscience is to drive the nervous system using periodic stimuli. This entrains neural responses at the stimulation frequency, resulting in high signal-to-noise ratios relative to single stimulus presentations. These periodic responses, often called the _steady-state_ or _frequency following_ response, can be recorded using invasive methods from single neurons [@Enroth-Cugell1966] and local field potentials [@Morrone1987], or with non-invasive electroencephalography (EEG) and magnetoencephalography (MEG) systems, both in humans [@Norcia2015] and in diverse animal species including insects [@Afsari2014], birds [@Porciatti1990], rodents [@Hwang2019] and primates [@Nakayama1982]. Steady-state methods are used to measure early sensory responses in vision [@Regan1966], hearing [@Rees1986] and somatosensation [@Snyder1992], and closely related paradigms have been developed to target specific stimulus features such as orientation [@Braddick1986], and facial expression [@Gray2020] and identity [@Liu-Shuang2014]. In fMRI research, _travelling wave_ methods [@Engel1994; @Sereno1995] are used to map the retinotopic responses of early visual cortex using stimuli that change periodically in spatial position. Finally, physiological reflexes such as the pupillary response to light can be entrained in a similar way [@Spitschan2014].

A convenient way to analyse the data from periodic stimulation experiments is to take the Fourier transform of the measured signal. The amplitude of the response at the stimulation frequency (and its harmonics - integer multiples of the stimulation frequency) is a precise and well-isolated index of the brain's response (see Figure \@ref(fig:fourierexplain)a,b). Fourier spectra comprise both amplitude and phase information that can be expressed in polar coordinates (Figure \@ref(fig:fourierexplain)c), or equivalently as complex numbers with real and imaginary components (Figure \@ref(fig:fourierexplain)d). In many studies the phase information is routinely discarded, and statistical comparisons are performed on the amplitude data (or equivalently power or signal-to-noise ratios) only [e.g. @Liu-Shuang2014; @Afsari2014; @Itthipuripat2014; @McFadden2014; @Vanegas2015; @Smith2017; @Hou2020] (note this is a non-exhaustive selection of recent studies). Typically researchers wish to test the experimental hypothesis that a signal is present against the null hypothesis that it is absent, or they wish to know if responses differ in amplitude and/or phase between two (or more) experimental conditions. However, there are several objections to using univariate statistics to answer these questions, as will be demonstrated below. An alternative is to use multivariate statistics, which take into account both the amplitude and phase information (represented as real and imaginary Fourier components in a Cartesian space). Multivariate methods have the advantage that they are more sensitive to weak signals, and therefore offer increased statistical power relative to univariate methods for typical applications (detecting the presence of a signal, or comparing multiple signals).

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/fourierexplain.pdf', bg="transparent", height = 8, width = 9)

@@

dev.off()}

knitr::include_graphics('figures/fourierexplain.pdf')

@@

\begin{equation}
\label{eq:t2c}
T^2_{circ} = (N-1)\frac{|\bar{x}-\mu|^2}{\Sigma|x_j - \bar{x}|^2},
\end{equation}

@@

In the present paper, best practice guidelines are developed for performing statistical tests on multivariate Fourier components derived from periodic stimulation paradigms. It is first demonstrated why parametric univariate statistics are inappropriate for such data, primarily because amplitudes for weak signals are not normally distributed. Then conditions are investigated under which either the $T^2$ or $T^2_{circ}$ statistic should be used. The range of sample sizes and effect sizes where $T^2_{circ}$ is more sensitive is identified. A novel method for testing the assumptions of the $T^2_{circ}$ statistic is developed, based on calculating the condition index of a multivariate data set. Appropriate methods for identifying outliers using the Mahalanobis distance are discussed, and a heuristic proposed. Next the logic of $T^2_{circ}$ is extended to situations with more than two levels of the independent variable, and the performance of this novel $ANOVA^2_{circ}$ statistic is compared to MANOVA. Finally, the proposed techniques are demonstrated on two example data sets (from mice and humans), and best practice guidelines are recommended for analysis decisions.

@@

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/amphists.pdf', bg="transparent", height = 6, width = 12)

@@

dev.off()}

knitr::include_graphics('figures/amphists.pdf')

@@

Typical solutions for dealing with skew, such as log-transforming the data, are unlikely to be equally applicable to all conditions. For example, if one wishes to compare a baseline where no stimulus was presented with a condition involving a strong signal, the former will be skewed and the latter normal. Applying a transform to both conditions is therefore problematic. Non-parametric statistics are a potential option, but these have generally lower statistical power than their parametric equivalents. Instead, the multivariate statistics discussed in the introduction avoid these issues and have greater statistical power, as will be demonstrated in the following section.

# Comparison of statistical power between univariate and multivariate tests

Using the $T^2$ or $T^2_{circ}$ statistic allows the phase information to be retained, and therefore provides greater power than running univariate T-tests on amplitudes, as well as avoiding problems caused by using absolute amplitude values. Victor & Mast [-@Victor1991] report simulations showing situations where $T^2_{circ}$ has greater power than $T^2$, which is expected on theoretical grounds because the additional degrees of freedom in the $T^2_{circ}$ test make it more efficient. Their simulations involved generating random data sets of different sample sizes, and different signal strengths (sample means), and comparing the number of such tests where each statistic was significant. The simulations showed the largest advantage for $T^2_{circ}$ for effect sizes around _d_ = 1 (where the sample mean is equal to the sample standard deviation). The advantage appeared to be stronger for smaller sample sizes.

Here these simulations are replicated and extended (see Figure \@ref(fig:powerfig)). The real and imaginary Fourier components of a periodic signal could in principle be analysed using two separate univariate t-tests. To detect the presence of a signal, each t-test would compare to a value of zero (i.e. the origin in Figure \@ref(fig:fourierexplain)d), to test against the null hypothesis that there is no signal ($H_{0x}: x = 0; H_{0y}: y = 0$). Without correction for familywise error, this approach has high statistical power for detecting when at least one test is significant (see grey triangles in Figure \@ref(fig:powerfig)a-c). However the use of two separate tests inflates the familywise error rate such that 10% of tests are significant (assuming $\alpha = 0.05$), even in the absence of a signal. Using Bonferroni correction to ameliorate this problem substantially reduces the power (white triangles in Figure \@ref(fig:powerfig)a-c).

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/powerfig.pdf', bg="transparent", height = 8, width = 12)

layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))

@@

tmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
tmatrix2 <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
tmatrix3 <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
tsqmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
circmatrix <- matrix(0,nrow=length(samplesizes),ncol=length(dlist))
for (n in 1:nsims){
  for (s in 1:length(samplesizes)){
    for (s2 in 1:length(dlist)){
      baseline <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
      data <- matrix(rnorm(2*samplesizes[s]),nrow=samplesizes[s],ncol=2)
      data[,1] <- data[,1] + dlist[s2]
      a <- tsqh.test(data)
      if (a[5]<0.05){tsqmatrix[s,s2] <- tsqmatrix[s,s2] + 1}
      a <- tsqc.test(data)
      if (a[5]<0.05){circmatrix[s,s2] <- circmatrix[s,s2] + 1}

      ampdata <- abs(complex(real = data[,1], imaginary = data[,2]))
      ampbase <- abs(complex(real = baseline[,1], imaginary = baseline[,2]))
      a <- t.test(ampbase,ampdata)
      if (a$p.value<0.05){tmatrix[s,s2] <- tmatrix[s,s2] + 1}

      a <- t.test(data[,1])
      b <- t.test(data[,2])
      if (a$p.value<0.05 | b$p.value<0.05){tmatrix2[s,s2] <- tmatrix2[s,s2] + 1}
      if (a$p.value<0.025 | b$p.value<0.025){tmatrix3[s,s2] <- tmatrix3[s,s2] + 1}

tpropM <- tmatrix/nsims
tpropM2 <- tmatrix2/nsims
tpropM3 <- tmatrix3/nsims
tsqpropM <- tsqmatrix/nsims
circpropM <- circmatrix/nsims
tsqvscirc <- circpropM - tsqpropM
tsqvscirc <- pmax(tsqvscirc,0)  # sets any negative values to 0

@@

for (n in 1:3){
  par(pty="s")
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
  axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
  axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
  mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)
  mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
  title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
  title(ylab="Proportion significant", col.lab=rgb(0,0,0), line=2.2, cex.lab=1.5)
  title(paste('d =',dlist[vinds[n]]),cex.main=2)
  lines(actualslog,tpropM3[,vinds[n]],lwd=3,col='black',lty=2)
  lines(actualslog,tpropM2[,vinds[n]],lwd=3,col='grey')
  lines(actualslog,tpropM[,vinds[n]],lwd=3,col=nbdtpal[3])
  lines(actualslog,tsqpropM[,vinds[n]],lwd=3,col=nbdtpal[1])
  lines(actualslog,circpropM[,vinds[n]],lwd=3,col=nbdtpal[2])

  points(actualslog,tpropM3[,vinds[n]],lwd=3,pch=25,cex=1.5,col='black',bg='white')
  points(actualslog,tpropM2[,vinds[n]],lwd=3,pch=24,cex=1.5,col='grey',bg='grey')
  points(actualslog,tpropM[,vinds[n]],lwd=3,pch=23,cex=1.5,col=nbdtpal[3],bg=nbdtpal[3])
  points(actualslog,tsqpropM[,vinds[n]],lwd=3,pch=21,cex=1.5,col=nbdtpal[1],bg=nbdtpal[1])
  points(actualslog,circpropM[,vinds[n]],lwd=3,pch=22,cex=1.5,col=nbdtpal[2],bg=nbdtpal[2])

  text(1.1,0.95,paste('(',letters[n],')',sep=''),cex=2)

  if (n==1){legend(1.4,1,c('Two univariate, uncorrected','Two univariate, corrected',expression(T^2),expression('T'[circ]^2), 'Univariate amplitudes'),pch=c(24:25,21:23),cex=1.5,pt.bg=c('grey','white',nbdtpal[1:3]),col=c('grey','black',nbdtpal[1:3]),box.lwd=2)}
}

par(pty="s")
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=c(1,5), ylim=c(0,4))
ticklocsx <- seq(1,5,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,4,1)    # locations of tick marks on y axis
ticklabelsx <- 2^ticklocsx        # set labels for x ticks
ticklabelsy <- seq(0,4,1)    # set labels for y ticks
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)  
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx, line=0.2)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Number of observations", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
title(ylab="Effect size (d)", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
title(expression(T^2),cex.main=2)
image(actualslog,dlist,tsqpropM,zlim=c(0,1),add=TRUE,col=kovesi.linear_blue_95_50_c20(256))
text(1.1,3.8,'(d)',cex=2)
legend(3.4,3.8,c('0','0.5','1'),lwd=6,cex=1.5,col=kovesi.linear_blue_95_50_c20(3),box.lwd=2,bg='white',title='Power')

@@

dev.off()}

@@

```{r powerfig, fig.cap="Simulations estimating the proportion of significant tests for simulated data with different sample sizes and effect sizes (100,000 simulated data sets per condition). Panels a-c replicate conditions reported by Victor and Mast (1991), and extend them to include univariate tests for comparison. Panels (d) and (e) show a wider range of conditions for each multivariate statistic. Panel (f) shows the difference between the two statistics, with contour lines indicating power differences of 0.02, 0.05, 0.1, 0.2 and 0.4.", fig.align="center", echo=FALSE, fig.width=12, fig.height=8}

knitr::include_graphics('figures/powerfig.pdf')

Using multivariate tests allows high power to be maintained without inflating the Type I error rate. Here, the multivariate mean (or equivalently, the difference in means between two conditions) is compared to zero, again testing against the null hypothesis that there is no signal ($H_0: x = y = 0$). The $T^2_{circ}$ statistic (red squares in Figure \@ref(fig:powerfig)a-c) has greater power than either the $T^2$ statistic (black circles) or the corrected univariate tests (white triangles). The power advantage over $T^2$ occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (when _d_ > 2 and N > 8). These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use can be restricted to small-sample studies.

As mentioned above, an alternative approach that is widely used in the literature is to discard the phase information and calculate the amplitude values instead (lengths of the lines in the top row of Figure \@ref(fig:amphists)). Because the amplitudes can never be negative, the presence of a signal must be detected by comparison to a baseline condition in which no stimulus was presented, typically using a paired t-test. Here, the null hypothesis is that the stimulus-present and stimulus-absent conditions produce equal amplitudes at the stimulus frequency ($H_0: A_{present} = A_{absent}$). This approach has markedly lower power than the multivariate tests (see blue diamonds in Figure \@ref(fig:powerfig)a-c). In principle, one could also conduct circular statistical tests on the phase component of the signal [@Berens2009]. However in the limit, testing both the amplitude and phase separately can be no more sensitive than testing the real and imaginary components (because these are just the polar and Cartesian representations of the same information), and suffers the same issues with familywise error from multiple tests. Multivariate approaches avoid these problems and provide high statistical power.

@@

Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on satisfying the $T^2_{circ}$ test's more stringent assumptions. The two dependent variables must be independent (i.e. uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (horizontal dashed lines). However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the sample variances of the two dependent variables grow more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2. In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/falsealarms.pdf', bg="transparent", height = 7, width = 6)

@@

dev.off()}

knitr::include_graphics('figures/falsealarms.pdf')

@@

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/distcomparison.pdf', bg="transparent", height = 5, width = 9)

@@

dev.off()}

knitr::include_graphics('figures/distcomparison.pdf')

@@

If a data set produces a significant result using the condition index test, this could be due to the presence of one or more outliers. The Mahalanobis distance [@Mahalanobis1936] is a useful metric for identifying such multivariate outliers so that they can be excluded. Note that outlier exclusion is contentious, but it is beyond the scope of this paper to contribute to this debate beyond noting that preregistration of any outlier exclusion protocols is adviseable. The Mahalanobis distance calculates the Euclidean distance between each data point and the sample mean, and scales it by the variance in the direction of the vector that joins the two points. This means that any correlations in the data set are taken into account when calculating the distance metric, _D_.

@@

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/outlierplot.pdf', bg="transparent", height = 6, width = 7)

@@

dev.off()}

knitr::include_graphics('figures/outlierplot.pdf')

@@

where $x_{i,k}$ is the $i$th data point from group _k_, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as $df_R = 2((Nk)-k)$, where _N_ is the number of observations per group, and _k_ is the number of groups. A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom. Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.

@@

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/powerfig2.pdf', bg="transparent", height = 8, width = 12)

layout(matrix(1:6,nrow=2,ncol=3,byrow=TRUE))

@@

text(1.1,0.95,paste('(',letters[n],')',sep=''),cex=2)

@@

dev.off()}

knitr::include_graphics('figures/powerfig2.pdf')

@@

```{r include=FALSE}

if (buildfigs==1){

pdf('figures/flowchart.pdf', bg="transparent", height = 4, width = 6)

par(mar=c(0.1,0.1,0.1,0.1))  
pos <- coordinates(c(1,2,4))
plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,1), ylim=c(0,1))

@@

dev.off()}

knitr::include_graphics('figures/flowchart.pdf')

@@

```{r include=FALSE}

@@

if (buildfigs==1){

pdf('figures/mousedata.pdf', bg="transparent", height = 6, width = 10)

@@

knitr::include_graphics('figures/mousedata.pdf')

@@

for (cond in 1:7){
  xvals <- Re(cleandata[,cond])/5
  yvals <- Im(cleandata[,cond])/5
  ellRot <- getel(data.frame(xvals,yvals))
  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)
  text(cond,1.02,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
}

dev.off()}
